<!-- LTeX: language=en-US -->
# Linear Regression {#sec-linearRegCh}


<!-- 
Points for future Revision: 

* Adding cluster robust SEs

 -->

::: {.hidden}
$$
\require{color}
%% Colorbox within equation-environments:
\newcommand{\highlight}[2][yellow]{\mathchoice%
  {\colorbox{#1}{$\displaystyle#2$}}%
  {\colorbox{#1}{$\displaystyle#2$}}%
  {\colorbox{#1}{$\displaystyle#2$}}%
  }%
$$
:::


This chapter is based on the following references:

* Chapter 3 of [An Introduction to Statistical Learning](https://www.statlearning.com/)
* Chapter 1 of Econometrics by Fumio Hayashi


## Motivation: Does Education Increase Wages? {#sec-motivation}

How much does one additional year of schooling increase a worker's hourly earnings? This question sits at the heart of labor economics and allows us to learn a lot what data can actually tell us. We observe in data that workers with more schooling tend to earn more — but how large is this association, how precisely can we estimate it, and can we say that education *causes* higher wages?

To answer these questions we use the `wage1` dataset from the `wooldridge` package, which contains data on 526 workers drawn from the 1976 Current Population Survey. @tbl-wage1-summary summarizes the key variables.

```{r}
#| label: tbl-wage1-summary
#| tbl-cap: "Key variables in the `wage1` dataset ($n = 526$ workers, 1976 Current Population Survey)."
library("wooldridge")
library("kableExtra")
data("wage1")

vars <- c("wage", "educ", "exper", "tenure", "female")
desc_tab <- data.frame(
  Variable    = vars, 
  Description = c("Hourly wage (USD)",
                  "Years of education",
                  "Years of work experience",
                  "Years with current employer",
                  "= 1 if female, 0 if male"),
  Mean = round(sapply(vars, \(v) mean(wage1[[v]])), 2),
  SD   = round(sapply(vars, \(v) sd(wage1[[v]])),   2),
  Min  = round(sapply(vars, \(v) min(wage1[[v]])),  2),
  Max  = round(sapply(vars, \(v) max(wage1[[v]])),  2),
  row.names = NULL
)
kbl(desc_tab,
    col.names = c("Variable", "Description", "Mean", "SD", "Min", "Max"),
    align = c("l", "l", "r", "r", "r", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The scatter plot below shows hourly wages (`wage`) against years of education (`educ`) for all $n=526$ workers. Because education is measured in integer years, we add a small horizontal jitter to avoid overplotting. The line is a preview of what this chapter builds up to: the **fitted simple linear regression** of wages on education.

```{r}
#| label: fig-wage-educ
#| fig-cap: >
#|   Hourly wages and years of education for $n=526$ workers.
#|   The line is the fitted simple linear regression; the shaded band
#|   is a 95% confidence interval for the regression line.
lm_simple <- lm(wage ~ educ, data = wage1)
x_seq     <- seq(min(wage1$educ), max(wage1$educ), length.out = 200)
ci        <- predict(lm_simple,
                     newdata  = data.frame(educ = x_seq),
                     interval = "confidence")

plot(jitter(wage1$educ, factor = 0.6), wage1$wage,
     xlab = "Education (years)",
     ylab = "Hourly Wage (USD)",
     pch  = 16,
     col  = adjustcolor("steelblue", alpha.f = 0.4),
     las  = 1, 
     ylim = range(wage1$wage, -3))
polygon(c(x_seq, rev(x_seq)),
        c(ci[, "lwr"], rev(ci[, "upr"])),
        col    = adjustcolor("lightcoral", alpha.f = 0.3),
        border = NA)
abline(lm_simple, col = "darkred", lwd = 2)
```

Two features of @fig-wage-educ will guide the entire chapter. First, the positive slope of the fitted line tells us that wages rise with education on average — but by exactly how much, and how sure are we? Second, the wide vertical spread of points around the line shows that education alone explains only part of the variation in wages: workers with the same years of schooling earn very different amounts. This raises the question of what else matters — experience, tenure, gender?

::: {.callout-note appearance="simple" icon="false"}
## Guiding Questions

1. **Estimation:** By how much does an additional year of education increase wages, on average?
2. **Inference:** Is this association statistically reliable, or could it reflect chance variation in our sample of $n=526$ workers?

We address a third, harder question in @sec-whatdoesbeta: does education *cause* higher wages, or do we merely observe an association?
:::

This chapter develops the statistical tools to answer questions 1 and 2 systematically. We begin with the **simple linear regression model** — wages regressed on education alone — and then extend to the **multiple linear regression model**, which lets us control for further variables such as experience, tenure, and gender simultaneously.


## Simple Linear Regression {#sec-simpleLinReg}

We begin with the simplest version of the linear regression model: a single
predictor $X_i$ explaining variation in an outcome $Y_i$. In the wages
example, $Y_i$ is the hourly wage and $X_i$ is the years of education of
worker $i$.

### The Model {#sec-simpleLR-model}

The **simple linear regression model** is given by

$$
Y_i = \beta_1 + \beta_2 X_i + \varepsilon_i, \quad i = 1, \dots, n,
$${#eq-simpleLR}

where:

* $Y_i$ is the **dependent variable** (hourly wage of worker $i$)
* $X_i$ is the **independent variable** (years of education of worker $i$)
* $\beta_1$ is the **intercept** — the expected wage when $X_i = 0$
* $\beta_2$ is the **slope** — the expected change in wages for one additional
  year of education
* $\varepsilon_i$ is the **"error" term** — capturing all other determinants of
  wages beyond education (ability, family background, occupation, parents, ...)


::: {.callout-note appearance="simple"}
## Crucial Assumption on $\varepsilon_i$: Exogeneity {-}

$$
E(\varepsilon_i \mid X_i) = 0, \quad i = 1, \dots, n
$$

That is, the mean of $\varepsilon_i$ is zero regardless of the value of $X_i,$
$$
E(\varepsilon_i \mid X_i = x) = 0, \quad i = 1, \dots, n
$$
for all possible values of $x.$ One often says: "$\varepsilon_i$ must be **mean independent** of $X_i.$" 

The mean independece or exogeneity assumption effectively requires that, conditional on education, no other factor drives wages that is also correlated with education. (Admittedly, this requirement is not readily visible from the above conditional mean statement, but one can show that this requirement is implied by the assumption.) 

**Problem:** Generally, we cannot proof that the exogeneity assumption is true, but need to argue its plausibility using specific expert knowledge about the data problem of interest. 
:::

The **simple linear regression model** @eq-simpleLR specifies the conditional mean of $Y_i$ given $X_i=x$ as a linear function
\begin{align*}
m(x)
& = E(Y_i \mid X_i = x) \\
& = E(\beta_1 + \beta_2 X_i + \varepsilon_i \mid X_i = x)  \\
& = E(\beta_1 + \beta_2 X_i\mid X_i = x) + \underbrace{E(\varepsilon_i \mid X_i = x)}_{=0}  \\
& = \beta_1 + \beta_2 x 
\end{align*}


The slope $\beta_2$ equals the partial derivative of the conditional mean function $m(x)=E(Y_i\mid X_i=x)$ with respect to $X_i=x$:

\begin{align*}
\frac{\partial}{\partial x} m(x) 
& = \frac{\partial}{\partial x} E(Y_i \mid X_i = x) \\
& = \frac{\partial}{\partial x} \left(\beta_1 + \beta_2 x\right) \\ 
& = \beta_2.
\end{align*}

In words: a one-unit increase in $X_i$ changes the *expected* value of $Y_i$
by $\beta_2$ units. In the wages example, $\beta_2$ measures the expected
wage premium per additional year of schooling, in dollars per hour.

::: {.callout-note appearance="simple"}
The model specifies the *conditional mean* of wages as linear in education.
It does not say every worker with the same education earns the same wage.
The error term $\varepsilon_i$ captures the deviation of worker $i$'s actual
wage from the conditional mean function $E(Y_i \mid X_i=x)=m(x).$
:::

### OLS Estimation {#sec-OLS-simple}

The parameters $\beta_1$ and $\beta_2$ are unknown. The **Ordinary Least
Squares (OLS)** estimator chooses $\hat\beta_1$ and $\hat\beta_2$ to minimize
the **Residual Sum of Squares (RSS)**:
$$
\operatorname{RSS} = \operatorname{RSS}(b_1,b_2) = \sum_{i=1}^{n} \bigl(Y_i - b_1 - b_2 X_i\bigr)^2.
$$

$$
(\hat\beta_1,\, \hat\beta_2)
= \arg\min_{b_1,\,b_2}
  \sum_{i=1}^{n} \bigl(Y_i - b_1 - b_2 X_i\bigr)^2.
$${#eq-OLS-min}

Taking partial derivatives of $\operatorname{RSS}(\beta_1,\beta_2)$ with respect to $\beta_1$ and $\beta_2$, setting them to zero, and solving yields the closed-form OLS estimators:

$$
\hat\beta_2
= \frac{\displaystyle\frac{1}{n}\sum_{i=1}^{n}(X_i - \bar X)(Y_i - \bar Y)}
       {\displaystyle\frac{1}{n}\sum_{i=1}^{n}(X_i - \bar X)^2}
= \frac{\widehat{\operatorname{Cov}}(X,\,Y)}{\widehat{\operatorname{Var}}(X)}
$${#eq-OLS-slope}

and

$$
\displaystyle \hat\beta_1 = \bar Y - \hat\beta_2\,\bar X,
$${#eq-OLS-intercept}

where $\bar X = \tfrac{1}{n}\sum_{i=1}^n X_i$ and
$\bar Y = \tfrac{1}{n}\sum_{i=1}^n Y_i$.

In R, the `lm()` function computes the OLS estimates:

```{r}
lm_simple <- lm(wage ~ educ, data = wage1)
coef(lm_simple)
```

We can verify @eq-OLS-slope directly:

```{r}
beta2_hat <- cov(wage1$educ, wage1$wage) / var(wage1$educ)
beta1_hat <- mean(wage1$wage) - beta2_hat * mean(wage1$educ)
c(beta1_hat = beta1_hat, beta2_hat = beta2_hat)
```

**Interpretation:** $\hat\beta_2 \approx 0.54$ means that on average each additional year
of education is associated with an increase of approximately 54 cents per hour
in wages. The intercept $\hat\beta_1 \approx -0.90$ is the predicted wage for a
worker with zero years of education. @fig-wage-educ shows that this predicted value is outside the observed wages $Y_i$ for $X_i=$`educ`$_i\approx 0.$ This issue means that our simple linear regression model is misspecified - a well specified model has its regression line in the center of the data cloud for *all* values of $X_i.$ Below, we will improve the model specification using a log-transformation $\log(Y_i)$; see @sec-logtrans.



### Assumptions {#sec-simpleLR-assumptions}

The OLS estimators $\hat\beta_1$ and $\hat\beta_2$ have desirable statistical
properties under four assumptions.

#### Assumption 1: Linear Model and Random Sampling {-}

The data $(Y_1, X_1), \dots, (Y_n, X_n)$ are a realization of an **i.i.d.
random sample** from the simple linear regression model in @eq-simpleLR. Random sampling means the $n$ observations are drawn independently from the same population distribution.

#### Assumption 2: Exogeneity {-}

$$E(\varepsilon_i \mid X_i) = 0.$$

The expected error is zero regardless of the value of $X_i$. This requires
that, conditional on education $X_i,$ no systematic factor drives wages that is also correlated with education $X_i.$

::: {.callout-caution appearance="simple"}
## When Does Exogeneity Fail?

Exogeneity fails when an omitted variable, say $W_i,$ is correlated with both $X_i$ and $Y_i$. 

Let, for instance, denote workers' ability by $W_i.$ This variable is hardly measurable and quantifiable and is thus usually contained in the error term $\varepsilon_i.$  

If higher-ability workers tend to acquire more education, i.e. $\operatorname{Cov}(X_i,W_i)>0,$ *and* if higher-ability workers tend to have higher wages, i.e. $\operatorname{Cov}(Y_i,W_i)>0,$ then $X_i$ is not exogenous and, equivalently, $\varepsilon_i$ is not mean-independent of $X_i$,
$$
E(\varepsilon_i \mid X_i) \neq 0.
$$ 
This is the **omitted variable problem**, revisited in @sec-OVB. In such a situation, we cannot trust our estimates. 
:::

#### Assumption 3: Variation in $X$ {-}

$$\operatorname{Var}(X_i) > 0.$$

The predictor must have positive variance. If all workers had identical
education levels, the denominator in @eq-OLS-slope would be zero and
$\hat\beta_2$ could not be computed.

#### Assumption 4: Homoskedasticity {-}

$$\operatorname{Var}(\varepsilon_i \mid X_i) = \sigma^2
  \quad\text{for all } i.$$

The variance of the error is constant across all values of $X_i$. The
scatter plot @fig-wage-educ suggests this may be violated: the spread of
wages around the regression line appears wider for more-educated workers.
We address this when introducing **heteroskedasticity-robust standard
errors** in @sec-robustSE.

### Properties of the OLS Estimator {#sec-simpleLR-props}

Under Assumptions 1–3, the OLS estimators are **unbiased**:

$$
E(\hat\beta_1) = \beta_1 \qquad \text{and} \qquad E(\hat\beta_2) = \beta_2.
$$

On average over many hypothetical random samples, OLS recovers the true
parameter values.

Under Assumptions 1–4, the variance of $\hat\beta_2$ conditional on
$(X_1,\dots,X_n)$ is

$$
\operatorname{Var}(\hat\beta_2 \mid X)
= \frac{\sigma^2}{\displaystyle\sum_{i=1}^{n}(X_i - \bar X)^2}.
$${#eq-var-beta1}

Three implications stand out:

* **More variation in $X$** (larger denominator) $\Rightarrow$ smaller
  variance $\Rightarrow$ more precise estimate.
* **Larger error variance $\sigma^2$** $\Rightarrow$ larger variance
  $\Rightarrow$ less precise estimate.
* **Larger sample size $n$** $\Rightarrow$ larger denominator
  $\Rightarrow$ more precise estimate.

Since $\sigma^2$ is unknown, we replace it by $\hat\sigma^2 = \text{RSE}^2$
to obtain the **estimated standard error**:

$$
\widehat{\operatorname{SE}}(\hat\beta_2)
= \frac{\text{RSE}}{\sqrt{\displaystyle\sum_{i=1}^{n}(X_i - \bar X)^2}}.
$${#eq-SE-beta1}

::: {.callout-note appearance="simple"}
## Gauss–Markov Theorem
Under Assumptions 1–4, OLS is the **B**est **L**inear **U**nbiased
**E**stimator (BLUE): among all unbiased estimators of $\beta_2$ that are
linear in $Y_1, \dots, Y_n$, OLS has the smallest variance.
:::

### Statistical Inference: $t$-Test for $\beta_2$ {#sec-inference-simple}

We have $\hat\beta_2 \approx 0.54$. But could this estimate simply reflect
chance variation in our sample of 526 workers, even if the true $\beta_2 = 0$?

**Connecting to Course 1.** You tested $H_0\colon \mu = \mu_0$ using

$$t = \frac{\bar Y - \mu_0}{s/\sqrt{n}} \sim t_{n-1}
  \quad\text{under } H_0.$$

The logic here is identical. We test

$$
H_0\colon \beta_2 = 0
\quad\text{(education has no effect on wages)}
$$

against the two-sided alternative $H_1\colon \beta_2 \neq 0$ using

$$
t = \frac{\hat\beta_2}{\widehat{\operatorname{SE}}(\hat\beta_2)}
  \sim t_{n-2}
  \quad\text{under } H_0 \text{ and Assumptions 1–4.}
$${#eq-t-stat}

The distribution is $t_{n-2}$ — not $t_{n-1}$ — because two parameters
have been estimated. We reject $H_0$ at significance level $\alpha$ when
$|t| > t_{n-2,\, 1-\alpha/2}$.

A $95\%$ **confidence interval** for $\beta_2$ is

$$
\hat\beta_2 \;\pm\; t_{n-2,\,0.975}
\cdot \widehat{\operatorname{SE}}(\hat\beta_2).
$${#eq-CI-beta1}

The `summary()` output reports $\hat\beta_j$,
$\widehat{\operatorname{SE}}(\hat\beta_j)$, the $t$-statistic, and the
$p$-value for each coefficient. `confint()` returns confidence intervals:

```{r}
summary(lm_simple)
confint(lm_simple, level = 0.95)
```

**Interpretation:** The $t$-statistic on `educ` far exceeds the critical
value and the $p$-value is essentially zero, so we reject
$H_0\colon\beta_2 = 0$ at any conventional significance level. The 95\%
confidence interval — approximately $[0.45,\, 0.63]$ — is entirely above
zero, confirming that the positive association between education and wages
is statistically reliable.

::: {.callout-caution appearance="simple"}
Statistical significance does not imply practical importance. A small
$p$-value tells us the association is unlikely to be zero; it does not tell
us whether the magnitude $\hat\beta_2 \approx 0.54$ is large enough to
matter. Always report and interpret the estimated size of the effect, not
just the $p$-value.
:::


## Multiple Linear Regression {#sec-multipleLinReg}

Education alone explains only 16.5% of wage variation. Other factors —
work experience, job tenure, gender — clearly matter. Adding them
simultaneously serves two goals: it improves prediction accuracy, and, as
we will see, it corrects for bias in the estimate of the education effect.

### Why Simple Regression Is Not Enough: Omitted Variable Bias {#sec-OVB}

Compare the OLS estimate of the education coefficient in two models:

```{r}
lm_educ_exper <- lm(wage ~ educ + exper, data = wage1)
rbind(
  simple   = coef(lm_simple)["educ"],
  multiple = coef(lm_educ_exper)["educ"]
)
```

The coefficient on `educ` changes when we add `exper`. This is not a
coincidence — it is **omitted variable bias (OVB)**.

**The OVB formula.** Suppose the true model contains two predictors:

$$
Y_i = \beta_1 + \beta_2 X_{1i} + \beta_3 X_{2i} + \varepsilon_i,
$${#eq-true-model}

but we estimate the misspecified simple regression
$Y_i = \alpha_0 + \alpha_1 X_{1i} + u_i$. The OLS estimate satisfies

$$
E(\hat\alpha_1) = \beta_2 + \underbrace{\beta_3 \cdot \hat\delta_1}_{\text{bias}},
$${#eq-OVB}

where $\hat\delta_1$ is the OLS slope from regressing $X_2$ on $X_1$.
The bias is non-zero whenever (i) $\beta_3 \neq 0$ (the omitted variable
affects $Y$) and (ii) $\hat\delta_1 \neq 0$ (it is correlated with $X_1$).

In the wages example, $X_1 = \text{educ}$ and $X_2 = \text{exper}$.
Workers who spent more years in school have fewer years of work experience,
so $\hat\delta_1 < 0$. Experience raises wages ($\beta_3 > 0$). The bias
$\beta_3 \cdot \hat\delta_1 < 0$ means the simple regression
*overstates* the return to education:

```{r}
delta1 <- coef(lm(exper ~ educ, data = wage1))["educ"]
beta2  <- coef(lm_educ_exper)["exper"]
c(delta1 = delta1, beta2 = beta2, bias = beta2 * delta1)
```

Adding `exper` also raises $R^2$, confirming that experience explains
additional variation in wages beyond what education captures:

```{r}
c(R2_simple   = summary(lm_simple)$r.squared,
  R2_multiple = summary(lm_educ_exper)$r.squared)
```

::: {.callout-caution appearance="simple"}
OVB is a **causal** problem, not a mechanical one. Including additional
control variables reduces bias only if those variables are themselves
exogenous. Adding a "bad control" — a variable that is affected by $X_1$
— can introduce new bias. We return to this in @sec-whatdoesbeta.
:::

### The Multiple Linear Regression Model {#sec-multipleLR-model}

The **multiple linear regression model** with $p$ predictors is

$$
Y_i = \beta_1 + \beta_2 X_{i1} + \dots + \beta_K X_{ip} + \varepsilon_i,
  \quad i = 1, \dots, n,
$${#eq-multipleLR}

or in matrix notation

$$
\underset{(n\times 1)}{Y}
= \underset{(n\times K)}{X}\;
  \underset{(K\times 1)}{\beta}
+ \underset{(n\times 1)}{\varepsilon},
$${#eq-multipleLR-matrix}

where

$$
Y = \begin{pmatrix}Y_1\\\vdots\\Y_n\end{pmatrix},\quad
X = \begin{pmatrix}
  1 & X_{11} & \cdots & X_{1p}\\
  \vdots & \vdots & \ddots & \vdots\\
  1 & X_{n1} & \cdots & X_{np}
\end{pmatrix},\quad
\beta = \begin{pmatrix}\beta_1\\\vdots\\\beta_K\end{pmatrix},\quad
\varepsilon = \begin{pmatrix}\varepsilon_1\\\vdots\\\varepsilon_n\end{pmatrix}.
$$

**Interpretation of $\beta_k$.** Each slope is a **partial effect**:

$$
\beta_k = \frac{\partial}{\partial X_{ik}} E(Y_i \mid X_i),
$$

the expected change in $Y_i$ for a one-unit increase in $X_{ik}$ *holding
all other predictors constant*. This is the key advantage over simple
regression.

The four assumptions from @sec-simpleLR-assumptions generalize by replacing
$X_i \in \mathbb{R}$ with $X_i \in \mathbb{R}^{K}$ and adding:

3. **Rank condition:** $\operatorname{rank}(X) = K$ almost surely
   — no predictor is a perfect linear combination of the others.

### OLS in Matrix Form {#sec-OLS-matrix}

The OLS estimator minimizes
$\text{RSS} = (Y - X\beta)^\top(Y - X\beta)$. Setting the gradient
to zero gives the **normal equations** $X^\top X\,\hat\beta = X^\top Y$.
Under Assumption 3, $X^\top X$ is invertible and the unique solution is

$$
\hat\beta = (X^\top X)^{-1} X^\top Y.
$${#eq-OLS-matrix}

The **fitted values** are

$$
\hat Y = X\hat\beta = X(X^\top X)^{-1}X^\top Y =: HY,
$${#eq-fitted-H}

where $H = X(X^\top X)^{-1}X^\top$ is the $(n \times n)$ **projection
matrix** (hat matrix). The diagonal elements $h_{ii} = [H]_{ii}$ are
**leverage statistics**: large $h_{ii}$ means observation $i$ is far from
the center of the predictor space and has an outsized influence on
$\hat\beta$.

```{r}
lm_multiple <- lm(wage ~ educ + exper + tenure, data = wage1)

X        <- model.matrix(lm_multiple)
Y        <- wage1$wage
beta_hat <- solve(t(X) %*% X) %*% t(X) %*% Y
round(cbind(manual = beta_hat, lm = coef(lm_multiple)), 4)
```

**Interpretation:** Controlling for experience and tenure, each additional
year of education is associated with approximately
`r round(coef(lm_multiple)["educ"], 2)` dollars/hour higher wages.

### Assessing Model Fit: $R^2$, Adjusted $R^2$, and RSE {#sec-multipleLR-fit}

$R^2$ from @eq-Rsquared **never decreases** when a predictor is added,
even if that predictor is pure noise. Comparing $R^2$ across models with
different numbers of predictors is therefore misleading.

The **adjusted $R^2$** penalizes for additional parameters:

$$
\bar R^2 = 1 - \frac{\text{RSS}/(n-K)}{\text{TSS}/(n-1)}
         = 1 - (1 - R^2)\frac{n-1}{n-K}.
$${#eq-adjR2}

$\bar R^2$ increases only when the new predictor reduces RSS by more than
chance. It can be negative for very poor models.

```{r}
#| label: tbl-model-comparison
#| tbl-cap: "Coefficient estimates and fit statistics across nested models."
library("kableExtra")

m1 <- lm(wage ~ educ,                  data = wage1)
m2 <- lm(wage ~ educ + exper,          data = wage1)
m3 <- lm(wage ~ educ + exper + tenure, data = wage1)
mods <- list(m1, m2, m3)

nms <- c("(Intercept)", "educ", "exper", "tenure")
coef_tab <- do.call(cbind, lapply(mods, function(m) {
  b        <- setNames(rep(NA_real_, length(nms)), nms)
  b[names(coef(m))] <- round(coef(m), 3)
  b
}))
rownames(coef_tab) <- nms

fit_tab <- rbind(
  R2    = sapply(mods, \(m) round(summary(m)$r.squared,     3)),
  AdjR2 = sapply(mods, \(m) round(summary(m)$adj.r.squared, 3)),
  RSE   = sapply(mods, \(m) round(summary(m)$sigma,         3)),
  n     = sapply(mods, nobs)
)
tab <- rbind(coef_tab, fit_tab)
colnames(tab) <- c("(1)", "(2)", "(3)")

kbl(tab, na_string = "—", align = "r") |>
  kable_styling(bootstrap_options = c("striped", "hover")) |>
  pack_rows("Coefficients", 1, 4) |>
  pack_rows("Fit statistics", 5, 8)
```

### Properties and Robust Standard Errors {#sec-robustSE}

**Unbiasedness.** Under Assumptions 1–3, $E(\hat\beta \mid X) = \beta$.

**Variance.** Under Assumptions 1–4,

$$
\operatorname{Var}(\hat\beta \mid X) = \sigma^2 (X^\top X)^{-1}.
$${#eq-Var-beta-matrix}

**Heteroskedasticity.** If $\operatorname{Var}(\varepsilon_i \mid X_i)
= \sigma_i^2$ varies across observations, @eq-Var-beta-matrix is
incorrect. The true variance has the **sandwich** form:

$$
\operatorname{Var}(\hat\beta \mid X)
= (X^\top X)^{-1}
  \Bigl(\sum_{i=1}^n \sigma_i^2\, X_i X_i^\top\Bigr)
  (X^\top X)^{-1}.
$${#eq-sandwich}

The **HC3** estimator replaces the unknown $\sigma_i^2$ with
$\hat{\varepsilon}_i^2/(1-h_{ii})^2$, which provides the best finite-sample performance
[@Long_Ervin_2000]. The `sandwich` and `lmtest` packages implement this:

```{r}
library("sandwich")
library("lmtest")
coeftest(lm_multiple, vcov = vcovHC(lm_multiple, type = "HC3"))
```

::: {.callout-note appearance="simple"}
In applied work, **always report HC3 robust standard errors**. The
coefficient estimates $\hat\beta$ are unchanged; only standard errors,
$t$-statistics, and confidence intervals are affected.
:::

### Statistical Inference {#sec-multipleLR-inference}

#### $t$-Tests for Individual Coefficients {-}

The $t$-test from @sec-inference-simple extends directly. Under
$H_0\colon\beta_j = 0$:

$$
t_j = \frac{\hat\beta_j}{\widehat{\operatorname{SE}}(\hat\beta_j)}
    \sim t_{n-K},
$${#eq-t-multiple}

with $n - p - 1$ degrees of freedom (one for each estimated parameter,
including the intercept).

#### $F$-Test for Joint Significance {-}

To test whether a *group* of predictors jointly matters, use the
**$F$-test**. The most common version tests all slopes simultaneously:

$$
H_0\colon \beta_2 = \beta_3 = \dots = \beta_K = 0.
$$

The $F$-statistic compares explained and unexplained variation:

$$
F = \frac{\text{ESS}/(K-1)}{\text{RSS}/(n-K)}
  \sim F_{p,\;n-K}
  \quad\text{under }H_0\text{ and Assumptions 1–4.}
$${#eq-F-stat}

`summary()` reports the overall $F$-statistic. Nested model comparisons
can be tested with `anova()`:

```{r}
anova(m1, m3)
```

#### Power and Consistency {-}

The **power** of the $t$-test — the probability of rejecting $H_0$ when
$\beta_j \neq 0$ — increases with $|\beta_j|$, with $n$, and with
variation in $X_j$. As $n \to \infty$, power $\to 1$ for any fixed
$\beta_j \neq 0$: OLS is **consistent**, $\hat\beta_j
\xrightarrow{p} \beta_j$.

The simulation below illustrates power as a function of sample size:

```{r}
#| label: fig-power-curve
#| fig-cap: >
#|   Simulated power of the two-sided t-test (α = 0.05) for
#|   H₀: β₁ = 0 as a function of sample size. True slope β₁ = 0.5;
#|   2 000 Monte Carlo replications per point.
set.seed(42)
ns    <- c(20, 40, 60, 100, 150, 200, 400)
nsim  <- 2000
power <- sapply(ns, function(n) {
  mean(replicate(nsim, {
    x <- rnorm(n)
    y <- 0.5 * x + rnorm(n)
    abs(coef(summary(lm(y ~ x)))[2, "t value"]) > qt(0.975, df = n - 2)
  }))
})
plot(ns, power,
     type = "b", pch = 16,
     xlab = "Sample size n", ylab = "Power",
     ylim = c(0, 1), las = 1)
abline(h = 0.05, lty = 2, col = "grey60")
text(max(ns), 0.07, "α = 0.05", adj = 1, col = "grey40", cex = 0.8)
```

### A Binary Predictor: The Gender Wage Gap {#sec-genderGap}

The variable `female` is a **binary (dummy) variable**: it equals 1 for
female workers and 0 for male workers. Adding it to the model

$$
Y_i = \beta_1 + \beta_2\,\text{educ}_i + \beta_3\,\text{exper}_i
    + \beta_4\,\text{tenure}_i + \beta_5\,\text{female}_i + \varepsilon_i
$${#eq-gender-model}

gives two parallel regression lines — one for men, one for women:

* **Men** ($\text{female}_i = 0$): intercept $\beta_1$
* **Women** ($\text{female}_i = 1$): intercept $\beta_1 + \beta_5$

So $\beta_5$ is the **adjusted wage gap**: the average wage difference
between women and men with identical education, experience, and tenure.

```{r}
lm_gender <- lm(wage ~ educ + exper + tenure + female,
                data = wage1)
coeftest(lm_gender, vcov = vcovHC(lm_gender, type = "HC3"))
```

The scatter plot below shows the two fitted lines (evaluated at mean
`exper` and `tenure`):

```{r}
#| label: fig-gender-lines
#| fig-cap: >
#|   Fitted wage-education lines for men (blue) and women (red) from
#|   model (4). Parallel lines impose identical returns to education
#|   for both groups.
beta       <- coef(lm_gender)
exper_bar  <- mean(wage1$exper)
tenure_bar <- mean(wage1$tenure)
educ_seq   <- seq(0, 18, length.out = 200)

wage_men   <- beta["(Intercept)"] +
              beta["educ"]   * educ_seq   +
              beta["exper"]  * exper_bar  +
              beta["tenure"] * tenure_bar
wage_women <- wage_men + beta["female"]

pt_col <- ifelse(wage1$female == 1,
                 adjustcolor("tomato",    alpha.f = 0.35),
                 adjustcolor("steelblue", alpha.f = 0.35))
plot(jitter(wage1$educ, factor = 0.6), wage1$wage,
     col = pt_col, pch = 16,
     xlab = "Education (years)", ylab = "Hourly Wage (USD)",
     las  = 1)
lines(educ_seq, wage_men,   col = "steelblue", lwd = 2)
lines(educ_seq, wage_women, col = "tomato",    lwd = 2)
legend("topleft",
       legend = c("Men", "Women"),
       col    = c("steelblue", "tomato"),
       lwd    = 2, bty = "n")
```

Model (4) assumes equal returns to education for men and women (parallel
lines). We relax this restriction with an interaction term in
@sec-interactions.


## Extensions of the Linear Model {#sec-extensions}

The multiple linear regression model $Y = X\beta + \varepsilon$ is
**linear in the parameters** $\beta$, not necessarily in the original
variables. We can include non-linear transformations of the predictors or
products of predictors as columns of $X$ — the model remains linear and
all results from @sec-multipleLinReg continue to apply.

### Log Transformations: The Mincer Equation {#sec-logtrans}

#### Motivation: The Distribution of Wages {-}

Hourly wages are right-skewed: a few high earners pull the mean well above
the median. Taking logarithms compresses the right tail and produces a
more symmetric distribution.

```{r}
#| label: fig-wage-hist
#| fig-cap: >
#|   Distribution of hourly wages (left) and log hourly wages (right)
#|   for 526 workers. Taking logarithms reduces right-skewness from
#|   2.0 to 0.4 and brings the distribution closer to symmetry.
par(mfrow = c(1, 2))
hist(wage1$wage,
     breaks = 25, col = "steelblue", border = "white",
     main = "Wage (USD/hour)",
     xlab = "Wage", las = 1)
hist(wage1$lwage,
     breaks = 25, col = "steelblue", border = "white",
     main = "Log Wage",
     xlab = "log(Wage)", las = 1)
par(mfrow = c(1, 1))
```

#### Interpretation of Log-Level Models {-}

In a **log-level model** $\log(Y_i) = \beta_1 + \beta_2 X_i +
\varepsilon_i$, the slope $\beta_2$ has a percentage interpretation.
Using the approximation $\log(1 + r) \approx r$ for small $r$:

$$
\beta_2 \approx
\frac{E[\log(Y_i) \mid X_i = x+1] - E[\log(Y_i) \mid X_i = x]}{1}
\approx \frac{\Delta Y}{Y},
$${#eq-log-interp}

so a one-unit increase in $X_i$ is associated with approximately a
$100 \cdot \beta_2\%$ change in $Y_i$. For larger $\hat\beta_2$, the
exact percentage change is $100 \cdot (\exp(\hat\beta_2) - 1)\%$.

#### The Mincer Earnings Equation {-}

The standard specification in the economics of education is the
**Mincer equation** [@Mincer1974]:

$$
\log(\text{wage}_i)
= \beta_1 + \beta_2\,\text{educ}_i + \beta_3\,\text{exper}_i
+ \beta_4\,\text{exper}_i^2 + \varepsilon_i.
$${#eq-mincer}

Compared to the levels models in @sec-multipleLinReg, this specification:

1. Accommodates the right-skewed wage distribution.
2. Gives $\hat\beta_2$ a natural interpretation as the **proportional
   return to one year of schooling**.
3. Includes a quadratic in experience to capture diminishing returns —
   a case of polynomial regression, examined in @sec-polynomial.

```{r}
lm_mincer <- lm(lwage ~ educ + exper + expersq, data = wage1)
summary(lm_mincer)
```

**Interpretation of $\hat\beta_\text{educ}$:**

```{r}
b_educ <- coef(lm_mincer)["educ"]
cat("Approximate return:", round(100 * b_educ, 1), "% per year\n")
cat("Exact return:      ", round(100 * (exp(b_educ) - 1), 1), "% per year\n")
```

Each additional year of education is associated with approximately a 9%
increase in hourly wages — a much more interpretable and economically
meaningful number than the 54 cents per hour from the simple levels model.

The fitted Mincer experience profile (at mean education):

```{r}
#| label: fig-mincer-exper
#| fig-cap: >
#|   Fitted Mincer wage-experience profile at mean education (12.6 years).
#|   Wages are back-transformed to USD/hour for interpretability.
x_seq   <- seq(1, 51, length.out = 300)
b       <- coef(lm_mincer)
log_fit <- b["(Intercept)"] + b["educ"] * mean(wage1$educ) +
           b["exper"] * x_seq + b["expersq"] * x_seq^2

plot(x_seq, exp(log_fit),
     type = "l", col = "darkred", lwd = 2, las = 1,
     xlab = "Experience (years)", ylab = "Predicted Wage (USD/hour)")
```

::: {.callout-note appearance="simple"}
**A note on $R^2$ comparisons.** The Mincer model has $R^2 \approx 0.30$
while the levels model `lm(wage ~ educ + exper + expersq)` has
$R^2 \approx 0.27$. These numbers are **not directly comparable** because
the dependent variable differs ($\log(\text{wage})$ vs $\text{wage}$).
The $R^2$ measures the fraction of variance of the respective outcome
explained — not of wages in a common unit.
:::

### Polynomial Regression {#sec-polynomial}

The Mincer equation @eq-mincer already contains $\text{exper}^2$. This is
an instance of **polynomial regression**: including higher-order powers of
a predictor as additional columns of $X$. The model is still linear in the
parameters, so OLS applies unchanged.

The general **quadratic regression model** is

$$
Y_i = \beta_1 + \beta_2 X_i + \beta_3 X_i^2 + \varepsilon_i.
$${#eq-polynomial}

The **marginal effect** of $X_i$ is now a linear function of $X_i$
itself:

$$
\frac{\partial}{\partial X_i} E(Y_i \mid X_i) = \beta_2 + 2\beta_3 X_i.
$${#eq-marginal-poly}

For the experience-wage relationship, we expect $\beta_2 > 0$ (experience
raises log wages) and $\beta_3 < 0$ (at a decreasing rate). We fit this
model using log wages for consistency with @sec-logtrans:

```{r}
lm_poly <- lm(lwage ~ exper + expersq, data = wage1)
summary(lm_poly)
```

The fitted curve and the marginal effect:

```{r}
#| label: fig-poly-exper
#| fig-cap: >
#|   Fitted quadratic regression of log wages on experience (left) and
#|   the implied marginal effect of one additional year of experience
#|   (right), in log-wage units (≈ percentage per year).
b     <- coef(lm_poly)
y_fit <- b["(Intercept)"] + b["exper"] * x_seq + b["expersq"] * x_seq^2
me    <- b["exper"] + 2 * b["expersq"] * x_seq

par(mfrow = c(1, 2))
plot(wage1$exper, wage1$lwage,
     col  = adjustcolor("steelblue", alpha.f = 0.3),
     pch  = 16, las = 1,
     xlab = "Experience (years)", ylab = "Log Wage")
lines(x_seq, y_fit, col = "darkred", lwd = 2)

plot(x_seq, me,
     type = "l", col = "darkred", lwd = 2, las = 1,
     xlab = "Experience (years)",
     ylab = "Marginal effect (log wage / year)")
abline(h = 0, lty = 2, col = "grey60")
par(mfrow = c(1, 1))
```

**Interpretation:** The marginal effect @eq-marginal-poly is zero at
$X^* = -\hat\beta_2/(2\hat\beta_3)$, the experience level where log wages
peak:

```{r}
peak_exper <- -coef(lm_poly)["exper"] / (2 * coef(lm_poly)["expersq"])
cat("Log wage peaks at experience =", round(peak_exper, 1), "years\n")
```

The peak at approximately 28–29 years of experience is consistent with
the Mincer literature: workers' wages rise steeply early in their careers
and level off or decline slightly near retirement age.

### Interaction Effects: Do Returns to Education Differ by Gender? {#sec-interactions}

We now combine the log-wage specification with an **interaction term** to
test whether the proportional return to education differs between men and
women. Adding $\text{educ}_i \times \text{female}_i$ to the Mincer
equation gives

$$
\log(\text{wage}_i)
= \beta_1 + \beta_2\,\text{educ}_i + \beta_3\,\text{exper}_i
+ \beta_4\,\text{exper}_i^2 + \beta_5\,\text{female}_i
+ \beta_6\,(\text{educ}_i \times \text{female}_i) + \varepsilon_i.
$${#eq-interact-model}

The fitted conditional means separate by gender:

* **Men** ($\text{female}_i = 0$): slope on education is $\beta_2$.
* **Women** ($\text{female}_i = 1$): slope on education is
  $\beta_2 + \beta_6$.

The interaction coefficient $\beta_6$ is the **difference in the
proportional return to education** between women and men.

```{r}
lm_interact <- lm(
  lwage ~ educ + exper + expersq + female + educ:female,
  data = wage1
)
coeftest(lm_interact, vcov = vcovHC(lm_interact, type = "HC3"))
```

```{r}
b <- coef(lm_interact)
cat("Return to education, men:  ",
    round(100 * (exp(b["educ"]) - 1), 1), "% per year\n")
cat("Return to education, women:",
    round(100 * (exp(b["educ"] + b["educ:female"]) - 1), 1),
    "% per year\n")
```

The non-parallel fitted log-wage lines (at mean `exper`):

```{r}
#| label: fig-interact-lines
#| fig-cap: >
#|   Fitted log-wage lines for men (blue) and women (red) from the
#|   interaction model. Non-parallel lines reflect different proportional
#|   returns to education by gender.
b          <- coef(lm_interact)
exper_bar  <- mean(wage1$exper)
educ_seq   <- seq(0, 18, length.out = 200)
lw_men     <- b["(Intercept)"] +
              b["educ"]    * educ_seq +
              b["exper"]   * exper_bar +
              b["expersq"] * exper_bar^2
lw_women   <- (b["(Intercept)"] + b["female"]) +
              (b["educ"] + b["educ:female"]) * educ_seq +
              b["exper"]   * exper_bar +
              b["expersq"] * exper_bar^2

pt_col <- ifelse(wage1$female == 1,
                 adjustcolor("tomato",    alpha.f = 0.3),
                 adjustcolor("steelblue", alpha.f = 0.3))
plot(jitter(wage1$educ, factor = 0.6), wage1$lwage,
     col = pt_col, pch = 16, las = 1,
     xlab = "Education (years)", ylab = "Log Wage")
lines(educ_seq, lw_men,   col = "steelblue", lwd = 2)
lines(educ_seq, lw_women, col = "tomato",    lwd = 2)
legend("topleft",
       legend = c("Men", "Women"),
       col    = c("steelblue", "tomato"),
       lwd    = 2, bty = "n")
```

If $\hat\beta_6$ is statistically significant, women face a different
proportional wage return per year of schooling — a finding with direct
implications for the causal analysis in @sec-whatdoesbeta.


## Detecting Potential Problems {#sec-diagnostics}

Before interpreting or reporting a fitted model, we should check whether
the model assumptions are plausible for the data at hand. Three
residual-based diagnostic plots catch the most common violations.

We illustrate using the Mincer equation `lm_mincer` from @sec-logtrans.
The `plot()` method for `lm` objects produces these plots automatically:

```{r}
#| label: fig-diagnostics
#| fig-cap: >
#|   Residual diagnostic plots for the Mincer equation
#|   log(wage) ~ educ + exper + exper². Panels: (1) residuals vs fitted,
#|   (2) normal Q-Q, (3) scale-location.
par(mfrow = c(1, 3))
plot(lm_mincer, which = 1)
plot(lm_mincer, which = 2)
plot(lm_mincer, which = 3)
par(mfrow = c(1, 1))
```

#### Plot 1: Residuals vs Fitted Values {-}

**What to look for:** Residuals should scatter randomly around zero with
no discernible pattern. A systematic curve suggests the functional form
is misspecified (e.g., a missing quadratic term). A funnel shape — where
spread increases with fitted values — indicates **heteroskedasticity**.

**In our model:** The residuals show no strong curvature, suggesting the
log transformation and experience quadratic have captured the main
non-linearities. Some spread remains, consistent with the variation in
wages that the model does not explain.

#### Plot 2: Normal Q-Q {-}

**What to look for:** Points should fall close to the 45° line. Heavy
tails (points curving away at both ends) indicate non-normal errors.

Under Assumptions 1–4, OLS is unbiased and the $t$- and $F$-tests are
exactly valid only if $\varepsilon_i$ is normally distributed. For large
$n$, the **Central Limit Theorem** ensures approximate validity even
without normality — so departures from normality matter most in small
samples.

**In our model:** The Q-Q plot shows mild right-skewness in the
residuals. With $n = 526$, the CLT provides adequate coverage and the
inference results remain reliable.

#### Plot 3: Scale-Location {-}

**What to look for:** The square root of the absolute standardized
residuals should be roughly flat across fitted values. An upward slope
indicates that larger fitted values are associated with larger residuals
— **heteroskedasticity**.

**In our model:** The scale-location plot shows a modest upward trend,
suggesting some heteroskedasticity remains even after the log
transformation. This reinforces the recommendation from @sec-robustSE:
**always use HC3 robust standard errors**.

::: {.callout-note appearance="simple"}
Diagnostic plots reveal problems; they do not fix them. The appropriate
response depends on the violation: heteroskedasticity → robust SEs;
non-linearity → transform predictors or add polynomial terms; outliers →
investigate individual observations, consider robust regression.
:::


## Prediction vs. Inference {#sec-prediction}

Linear regression is used for two distinct purposes that require different
tools and different interpretations of uncertainty.

**Inference on parameters** answers: *By how much does education raise
wages on average?* The target is $\beta_2$, a population parameter. The
relevant measure of uncertainty is $\widehat{\operatorname{SE}}(\hat\beta_2)$.

**Prediction** answers: *What wage would we expect for a specific new
worker?* The target is $Y_\text{new}$ for a given $X_\text{new}$. The
relevant measure of uncertainty must account for two sources of error:
uncertainty about $\beta$ *and* the inherent randomness of $Y_\text{new}$
around its conditional mean.

### Confidence Interval for the Conditional Mean {-}

The first target is $E(Y_\text{new} \mid X_\text{new}) = X_\text{new}^\top\beta$.
Its estimator is $X_\text{new}^\top\hat\beta$, with variance

$$
\operatorname{Var}(X_\text{new}^\top\hat\beta \mid X)
= \sigma^2\, X_\text{new}^\top (X^\top X)^{-1} X_\text{new}.
$${#eq-CI-mean-var}

A $95\%$ **confidence interval for the conditional mean** is

$$
X_\text{new}^\top\hat\beta \;\pm\;
t_{n-K,\,0.975}\cdot
\sqrt{\hat\sigma^2\, X_\text{new}^\top(X^\top X)^{-1}X_\text{new}}.
$${#eq-CI-mean}

### Prediction Interval for a New Observation {-}

The second target is a *new individual outcome*
$Y_\text{new} = X_\text{new}^\top\beta + \varepsilon_\text{new}$.
Its prediction error is

$$
Y_\text{new} - X_\text{new}^\top\hat\beta
= \underbrace{X_\text{new}^\top(\beta - \hat\beta)}_{\text{estimation error}}
+ \underbrace{\varepsilon_\text{new}}_{\text{irreducible error}},
$$

so the variance of the prediction error is

$$
\operatorname{Var}(Y_\text{new} - X_\text{new}^\top\hat\beta \mid X)
= \sigma^2\bigl(1 + X_\text{new}^\top(X^\top X)^{-1}X_\text{new}\bigr).
$${#eq-PI-var}

The extra $\sigma^2$ in @eq-PI-var — absent in @eq-CI-mean-var — reflects
the irreducible randomness of the individual outcome. A $95\%$ **prediction
interval** is therefore always **wider** than the confidence interval for
the mean:

$$
X_\text{new}^\top\hat\beta \;\pm\;
t_{n-K,\,0.975}\cdot
\sqrt{\hat\sigma^2\bigl(1 + X_\text{new}^\top(X^\top X)^{-1}X_\text{new}\bigr)}.
$${#eq-PI}


### Assessing Model Fit: $R^2$ and RSE {-}

How well does education alone explain wages? We measure this through a
decomposition of the total variation in $Y$.

The **fitted value** $\hat Y_i = \hat\beta_1 + \hat\beta_2 X_i$ is the
model's prediction for worker $i$. The **residual**
$\hat{\varepsilon}_i = Y_i - \hat Y_i$ is the prediction error.

The total variation in $Y$ decomposes into explained and unexplained parts:

$$
\underbrace{\sum_{i=1}^{n}(Y_i - \bar Y)^2}_{\text{TSS}}
=
\underbrace{\sum_{i=1}^{n}(\hat Y_i - \bar Y)^2}_{\text{ESS}}
+
\underbrace{\sum_{i=1}^{n} \hat{\varepsilon}_i^2}_{\text{RSS}}
$${#eq-SS-decomp}

where TSS is the **Total Sum of Squares**, ESS is the **Explained Sum of
Squares**, and RSS is the **Residual Sum of Squares**.

The **coefficient of determination** $R^2$ is the fraction of total
variation explained by the model:

$$
R^2 = \frac{\text{ESS}}{\text{TSS}} = 1 - \frac{\text{RSS}}{\text{TSS}},
\qquad 0 \leq R^2 \leq 1.
$${#eq-Rsquared}

The **Residual Standard Error (RSE)** estimates the typical size of the
residuals:

$$
\text{RSE} = \sqrt{\frac{\text{RSS}}{n - 2}}.
$${#eq-RSE}

The denominator is $n-2$ because two parameters ($\hat\beta_1$,
$\hat\beta_2$) have been estimated. The RSE is in the same units as $Y$
(here, dollars per hour).

```{r}
RSS <- sum(residuals(lm_simple)^2)
TSS <- sum((wage1$wage - mean(wage1$wage))^2)
R2  <- 1 - RSS / TSS
RSE <- sqrt(RSS / (nrow(wage1) - 2))
c(R2 = R2, RSE = RSE)
```

These match the values reported by `summary()`:

```{r}
summary(lm_simple)$r.squared
summary(lm_simple)$sigma
```

**Interpretation:** $R^2 \approx 0.165$ means education explains about
16.5% of the variation in wages across workers. The RSE of approximately
\$3.38 per hour reflects how much individual wages deviate from the fitted
line on average. Both values confirm that education is informative about
wages — but that many other factors matter, motivating the extension to
multiple regression in @sec-multipleLinReg.



### Application to the Wages Example {-}

Consider a new worker with 16 years of education (college graduate) and
5 years of work experience. What wage do we predict?

```{r}
new_worker <- data.frame(educ = 16, exper = 5, expersq = 25)

ci <- predict(lm_mincer, newdata = new_worker, interval = "confidence")
pi <- predict(lm_mincer, newdata = new_worker, interval = "prediction")

cat("Point estimate (USD/hour):             ", round(exp(ci[,"fit"]), 2), "\n")
cat("95% CI for mean wage (USD/hour):  [",
    round(exp(ci[,"lwr"]), 2), ",", round(exp(ci[,"upr"]), 2), "]\n")
cat("95% PI for individual wage (USD/hour): [",
    round(exp(pi[,"lwr"]), 2), ",", round(exp(pi[,"upr"]), 2), "]\n")
```

The confidence interval [$5.43, $6.23] is narrow: we can estimate the
*average* wage of all workers with this profile quite precisely. The
prediction interval [$2.42, $14.01] is nearly six times as wide: predicting
*one individual's* wage is far less certain, because wages vary enormously
among workers with identical observable characteristics.

::: {.callout-note appearance="simple"}
As $n \to \infty$, the CI width $\to 0$ — we eventually learn $\beta$
exactly. The PI width does **not** shrink to zero: the irreducible
$\sigma^2$ term in @eq-PI-var remains even in infinite samples. No matter
how much data we have, an individual's wage contains randomness that no
model can eliminate.
:::

The figure below visualises both bands along the education dimension,
holding `exper` at its sample mean:

```{r}
#| label: fig-CI-PI
#| fig-cap: >
#|   95% confidence band for the conditional mean (dark red, narrow) and
#|   95% prediction band (salmon, wide) from the Mincer model, plotted
#|   against education at mean experience.
educ_seq  <- seq(0, 18, length.out = 200)
exper_bar <- mean(wage1$exper)
nd        <- data.frame(educ    = educ_seq,
                        exper   = exper_bar,
                        expersq = exper_bar^2)
ci_band <- predict(lm_mincer, newdata = nd, interval = "confidence")
pi_band <- predict(lm_mincer, newdata = nd, interval = "prediction")

plot(wage1$educ, wage1$lwage,
     col  = adjustcolor("steelblue", alpha.f = 0.3),
     pch  = 16, las = 1,
     xlab = "Education (years)", ylab = "Log Wage")
polygon(c(educ_seq, rev(educ_seq)),
        c(pi_band[,"lwr"], rev(pi_band[,"upr"])),
        col = adjustcolor("lightcoral", alpha.f = 0.25), border = NA)
polygon(c(educ_seq, rev(educ_seq)),
        c(ci_band[,"lwr"], rev(ci_band[,"upr"])),
        col = adjustcolor("darkred", alpha.f = 0.35), border = NA)
lines(educ_seq, ci_band[,"fit"], col = "darkred", lwd = 2)
legend("topleft",
       legend = c("Fitted mean", "95% CI (mean)", "95% PI (individual)"),
       col    = c("darkred",
                  adjustcolor("darkred",     alpha.f = 0.5),
                  adjustcolor("lightcoral",  alpha.f = 0.5)),
       lwd    = c(2, 8, 8), bty = "n")
```


## What Does $\hat\beta_2$ Really Tell Us? {#sec-whatdoesbeta}

We now have a complete regression toolkit. The Mincer equation from
@sec-logtrans estimates that each additional year of education is associated
with approximately a **9% increase in hourly wages**, controlling for
experience. The prediction machinery from @sec-prediction tells us this
estimate is statistically precise and the model fits the data reasonably
well ($R^2 \approx 0.30$).

So — does education make workers more productive and therefore better
paid? Can a student reading this script expect a 9% wage premium for each
additional year they spend in university?

**Not necessarily.** The OLS estimate $\hat\beta_\text{educ} \approx 0.09$
measures a *statistical association*, not a causal effect. To see why,
consider two scenarios that are observationally indistinguishable with
our data:

**Scenario A (education is causal):** More education raises productivity,
which raises wages. The 9% estimate reflects the true causal return.

**Scenario B (ability bias):** Some workers are more able — they learn
faster, work harder, think more creatively. High-ability workers both
*choose* to acquire more education (because it is less costly or more
rewarding for them) and *earn higher wages* (for reasons unrelated to
their education). In this scenario, the regression picks up the combined
effect of education *and* the underlying ability differences between
workers who chose different education levels.

Under Scenario B, $E(\varepsilon_i \mid \text{educ}_i) \neq 0$ — **Assumption
2 (exogeneity) is violated**. The omitted variable `ability` is correlated
with both `educ` (more able workers get more education) and `wage` (more
able workers earn more), so by the OVB formula @eq-OVB:

$$
E(\hat\beta_\text{educ}) = \beta_\text{educ}^\text{causal}
+ \underbrace{\beta_\text{ability} \cdot \hat\delta_\text{educ}}_{\text{ability bias} > 0}.
$$

The OLS estimate *overstates* the causal return to education. Studies
exploiting natural experiments — such as compulsory schooling laws that
forced some students to stay in school longer — typically find smaller
causal returns than naive OLS estimates [@Card1999].

**Prediction vs. causation.** Crucially, none of this undermines the
model's *predictive* usefulness. Even if $\hat\beta_\text{educ}$ is
biased as a causal estimate, the Mincer equation still produces good
predictions of wages for workers similar to those in the sample. The
confidence and prediction intervals in @sec-prediction remain valid.
Bias and prediction accuracy are distinct properties of an estimator.

::: {.callout-note appearance="simple"}
## What We Have and What We Are Missing

| | **Prediction** | **Causal inference** |
|---|---|---|
| Goal | Forecast $Y_\text{new}$ | Estimate $\partial Y / \partial X$ |
| Key assumption | Stable data-generating process | Exogeneity / no OVB |
| OLS sufficient? | Often yes | Only under strict conditions |
| Question answered | "What wage would this worker likely earn?" | "What wage *would* this worker earn with one more year of education?" |

Answering causal questions requires additional assumptions and methods
beyond OLS — the topic of the next chapter.
:::

The returns-to-education question has motivated decades of empirical
research and methodological innovation: instrumental variables,
regression discontinuity, difference-in-differences. We take up these
tools in @sec-causalInference.

